Separation of an upwelling current bounding the Juan de Fuca Eddy, off Vancouver Island

### 2022-04-09 PW13Paper.tex
New version

The southern Vancouver Island continental shelf is very productive due to high nutrient input from the Strait of Juan de Fuca and Salish Sea estuarine system, and substantial cross-shelf transport due to the complicated topography.  The estuarine water flows poleward, hugging the coast as the Vancouver Island current, and in the summer shelf water flows equatorward.  Trapped between these two currents is the Juan de Fuca eddy, a stagnant region of water that occupies the shelf where it abruptly widens just north of the Strait of Juan de Fuca.

@@

Here we present detailed observations from the southern Vancouver Island shelf, collected in summer 2013, and sampled with a rapid profiling vehicle equipped with a CTD and oxygen sensor, supplemented by traditional hydrographic surveys. The shelf has complicated bathymetry \fref{fig:LocMapBoth}, with a somewhat typical shelf poleward of our study site, but then widening in the lee of La Perouse Bank, and then hitting the Juan de Fuca canyon, equatorward.  Water from the strait of Juan de Fuca flows poleward as a buoyant current that hugs the coast \cite{thomsonetal89, hickeyetal91}, while shelf water flows from poleward in the summer under upwelling conditions, both forced by local winds and via tele-connections with the long, homogenous shelves equatorward off Washington and Oregon \cite{hickeyetal91,thomsonkrassovski15,engidaetal16}.  Trapped between these two currents is a region of relatively homogenous water that has been termed the Juan de Fuca eddy \cite{freelanddenman82,freelandmcintosh89,foremanetal07,macfadyenhickey10}.

One of our findings below is that the water in the Juan de Fuca eddy is apparently quite old.  The low oxygen found in the Juan de Fuca eddy has caused it to be inferred to be upwelled from as deep as 400 m \cite{freelanddenman82,deweycrawford88}.  This inference was based on the oxygen being conservative, and hence having to come from close to the oxygen minimum zone found further offshore \cite{mackasetal87}.  This lead to hypotheses that perhaps the eddy was fed by waters being drawn up a spur canyon of the Juan de Fuca canyon (often called the Tully Canyon) via ageostrophic transport from offshore to onshore due to the low pressure in the eddy center \cite{weaverhsieh87}.  Below, we will argue that there is not substantial evidence of such transport, and that oxygen is likely low because of consumption due to respiration.

A second important finding is that the water flowing equatorward along the shelf from upstream is all displaced from the shelf and pushed offshore and replaced by offshore water that floods the southern shelf.  Such events are important as they tend to transport nutrient rich shelf water out into the open ocean, fueling offshore ecosystems.
This cross-shelf exchange has not, to our knowledge, been remarked upon for the Vancouver Island shelf.  Mesoscale separations have been observed further to the south off Cape Blanco \cite{barthetal00}, and was also inferred at other locations along the coast.  Similar observations have been made elsewhere \cite[e.g.]{relvasbarton05}, though usually in the presence of a distinct coastal headland, a feature not found at our study location.  The mechanisms driving such separations are poorly understood, but hypotheses include coastal hydraulics \cite{dalebarth01} leading to along-shore trapping of coastal waves,  wind stress curl variations \cite{castelaobarth07}, and induced relative vorticity due to stretching of parcels that inertially overshoot their initial isobaths due to a sudden change in downstream bathymetry.

In this paper we present our observations from 2013 by describing our at-sea observation tools (\fref{sec:Site}), presenting the data in a number of ways to highlight the important processes, with a focus on the offshore side of the eddy region (\fref{sec:Observations}) where we consider the age of the eddy, the steadiness of the front separating the eddy and the offshore water, properties along the spur canyon that was believed to feed the eddy, and finally describe the large offshore separation of the shelf-break current.  We discuss the origin of the eddy and the implications of the separating jet.

@@

\begin{figure*}[htbp]
  \includegraphics[width=5in]{LocMapBoth.pdf}
  \caption{a) Study site on the Vancouver Island Shelf.  The blue arrow indicates the direction of the Vancouver Island Current.  The pink arrow indicates southward flow of the coastal upwelling current.  The dashed box indicates the approximate limits of the study area.  b) The study area with hydrographic casts from the La Perouse cruises along the LB and LC Lines are green dots, hydrographic casts during the Falkor cruise indicated in yellow dots, and Moving Vessel Profiler casts as red dots.  The coordinate system used for this paper is shown with alternating grey and white bands at 10-km intervals.}
  \label{fig:LocMapBoth}
\end{figure*}

At the south end of the study site is the Juan de Fuca canyon, which feeds dense water into Juan de Fuca Strait.  This dense water is mixed with fresh water from the Fraser River at the sills and archipelagos further inland, and fluxes out the Strait again, where it turns poleward along Vancouver Island to form the Vancouver Island Current.  The Juan de Fuca Canyon has a notable spur canyon, the Tully Canyon, that incises the shelf towards the north into the study site (\fref{fig:LocMapBoth}b).  The rest of the shelf is punctuated by a series of banks and shallow basins.  Notable is La Perouse bank, running along the shelf from the north, and moving isobaths offshore until it abruptly ends at around 48.5 N. The shelf break is characterized by a series of submarine canyons, in particular Nitnat Canyon and Barkley Canyon.

@@

Observations aboard the \emph{R/V Falkor} were carried out from 2013-08-21 to 2013-08-30, starting two weeks before the September Tully cruise.  Data were collected with an ODIM Brooke-Ocean Moving Vessel Profiler, profiling to 200 m.  The MVP was equipped with an AML Oceanographic CTD, and a Rinko Oxygen sensor with a 7-s response time foil.  The MVP profiles at vertical fall speeds of $3\ \mathrm{m\,s^{-1}}$ to depths of 200 m or to within 5 m of the seafloor, whichever is shallower.  Data collection took place while the ship cruised at speeds between 5 and 8 kts, usually at around 6 kts to enable fine horizontal spacing of the casts, with typical spacing less than 800 m.  Data is collected during down- and up-casts, but is only reported for down-casts because the path through the water is highly asymmetric on the up-cast, whereas the downcast is largely vertical.  The rapid speed of the profiling makes the oxygen measurements somewhat coarse, and probably biased due to the phase lag of the sensor, so we treat these largely qualitatively in this paper.

@@

Winds during the cruises were typical for the west coast of Vancouver Island, with down-welling favorable winds during July and early August, and then intermittent winds before and during the Falkor cruise.  Note that this locale is strongly affected by coastally trapped waves from further south, so upwelling usually persists despite local wind forcing \cite{thomsonkrassovski15, engidaetal16}.

@@

Hydrographic sections along the LB and LC lines highlight summer conditions on the Southern Vancouver Island Shelf, and indicate some of the features we are focusing on in this paper (\fref{fig:LaPerouse2013Ctd}).  During both cruises and at both lines there is clear evidence of upwelling, with the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal reaching from 130 m to shallower than 85 m over the shelf.  This dense water tends to be low in oxygen and cool. Further onshore, isopycnals tilt back down, and warm water fills the water column, evidence of the Vancouver Island Current that hugs the coast here.

@@

The spice anomaly defined this way is mapped in the middle row of each set of plots in \fref{fig:LaPerouse2013Ctd}.  Offshore water tends to be high is absolute spice anomaly, with a positive-anomaly layer (red) centered at $26.4\ \mathrm{kg\,m^{-3}}$ sandwiched between negative-anomaly layers above and below.  On the shelf ($-10\, \mathrm{km} < X < 30\, \mathrm{km}$), the distinct $\theta$--$S$ masses are attenuated and much closer to the mixing line than the offshore water.  Hugging the coast ($X > 30\, \mathrm{km}$), is the Vancouver Island Current, with a negative spice anomaly in June, and a positive anomaly in summer, consistent with the warming of this water during the summer.

@@

A systematic survey through the front between the offshore water and the ``eddy'' water demonstrates the structure of the front (\fref{fig:Frontsurvey}).  The survey started close to shore and passed through the Vancouver Island Current (casts 2030--2080).  This water is buoyant and signficiantly fresher and cooler than water at the same density near the surface.  There is strong evidence of isopycnal mixing of the surface water and the outflow water, with warm-salty water in the relatively wide front between the ``eddy'' water and the water in the current.

@@

\subsection{Tully Canyon}

The seasurface is low in the middle of the ``Eddy'', so water has been hypothesized to move up Tully Canyon due to ageostrophic motion \cite{weaverhsieh87,freelanddenman82}.  This is difficult to infer from water properties alone.  Three transects up the canyon indicate that deep isopycnals slope up into the canyon (\fref{fig:CanyonPropertiesSpice}, to approximately 30 km).  Across the eddy, isopycnals are largely flat until they intersect the Vancouver Island current ($x\approx 80\ \mathrm{km}$).  The deeper isopycnals do not make it into the deeper basin northeast of La Perouse bank ($x\approx 70\ \mathrm{km}$).

@@

Much of the modified water in the Tully Canyon appears to come from the shelf to the west, but heavily modified by tidal moxing. A repeated tidal survey over the bank on the west side of the canyon shows a strong hydraulic response (\fref{fig:HydraulicsCanyon}) during onshore flow (17:58--21:00).  Note the tide here is largely diurnal, so this onslope flow only occurs once a day.  Given the stratification of $N\approx 6\times10^{-3} \ \mathrm{rad\,s^{-1}}$ and an overturning scale of 50 m, we might expect dissipation rates reaching $\epsilon \sim L^2N^{3} \approx 5\times10^{-4}\ \mathrm{m^2\,s^{-3}}$, which is more than three orders of magnitude higher than dissipation observed west of this location by \cite{deweycrawford88}.  The effect on the properties in the canyon is for water to spill over from the offshore front into the canyon during the tide.  This water is rapidly mixed with surrounding water such that its strong offshore spice values are attenuated.

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=4in]{HydraulicsCanyon}
    \caption{Repeated, tide-resolving survey across the Tully Canyon, time is indicated in the upper left of each plot (29 August, 2013).
      \label{fig:HydraulicsCanyon}
    }
  \end{center}
\end{figure*}

@@

We can attempt to track this feature in satellite imagery, though there are lots of clouds during this time period.  Sea-surface temperatures are generally colder coming out of the Strait of Juan de Fuca (\fref{fig:SST}).  Water flowing equatorward along the shelf tends to be cooler that offshore water, though this water is significantly saltier than the Juan de Fuca water hugging the shelf. On 2013-08-26, there is a cold tongue of water separating from La Perouse bank, and pointing south at 125.8 W.  Further downstream, there is the hint of a slightly cooler tendril streaming west at 48 N.  This feature is not as well-developed in the previous image, 2013-08-20, perhaps indicating that it is an evolving mesoscale feature (or a consequence of the cloud cover making for smeared 1-d composites)

\begin{figure}
  \begin{center}
    \includegraphics[width=4in]{SST}
    \caption{Sea surface temperature composites \cite{GHRSST, ChinEtAl17} from the observation period.  Note these composites are during a time of year when there is substantial cloud cover.\label{fig:SST}}
  \end{center}
\end{figure}

A similar impression is given by ocean chlorophyll estimates (\fref{fig:ChlA}).  The 8-d composites show a region of high chlorophyll  on the west side of the eddy. It also shows a relatively  high-chlorophyll squirt to the west, again exiting the study region at approximately 48 degrees N.  The feature is relatively long-lived, on the order of one month.  Inspection of images before 2013-08-05 were too obscured by clouds, or did not show this feature, and even by 2013-09-06 we see the feature fading from the satellite.  However, note that this feature is somewhat south of tongue that we observe deeper in the water column, but is indicative of substantial offshore transport in the region.  On the scale of mesoscale features along the coast it is not unusually large, but is a facet of the local oceanography not often commented on.

@@

Unfortunately, this data set cannot definitively track where the water in the eddy originated from.  The deepest water in the eddy could originate along the $\theta$--$S$ line from approximately 5.5 degrees to 7.5 degrees (\fref{fig:LaPerouse2013TS}a), which in the open ocean spans depths from 420 m to 70 m.  \cite{mackasetal87} attempted to determine the origin of the water by including oxygen as a third variable to resolve the ambiguous $\theta$--$S$ relation.  However, as \fref{fig:LaPerouse2013TS}b makes clear, oxygen is definitely not a conserved property in the eddy, with concentrations up to $150 \ \mathrm{\mu mol\ kg^{1}}$ lower in the eddy than the water found offshore, and in a way that definitely cannot result from conservative mixing.  As a best guess, if the water in the deepest part of the eddy came from a vertical mixture of the water at $26.5\,\mathrm{kg\,m^{-3}}$ and an equal distance down in density space of $26.7\,\mathrm{kg\,m^{-3}}$, then the deepest water in the eddy would be coming from a depth of 250 m.  This is actually a relatively typical upwelling depth for coastal flows, and may not require extra input up Tully canyon as posited by \cite{freelanddenman82}.

We found little evidence of vigorous flow up the Tully Canyon, or if there is, then the mixing in the canyon is quick enough to remove the $\theta$--$S$ signature of offshore water within 20 km (\fref{fig:CanyonPropertiesSpice}).  We did find substantial evidence of mixing in the canyon, but the primary route of high-spice water into the canyon appears to be due to tidal flow over the banks on the west side (\fref{fig:HydraulicsCanyon}), rather than evidence of flow up the canyon.

There is evidence of aging of the water in the eddy between late spring and late summer (\fref{fig:LaPerouse2013Ctd}), with a reduction of oxygen in the eddy.  If we posited that the reduction was all from the same water, then the oxygen consumption rate over the time between the late-May and early September cruises would be on the order of $0.5\ \mathrm{\mu mol\, kg^{-1} d^{-1}}$.  This seems on the low side of continental upwellng system estimates of apparent oxygen utilization rates, which are between $1$ and $5\  \mathrm{\mu mol\, kg^{-1} d^{-1}}$ \cite{dortchetal94}. \marginpar{More citations?  Is AOU really poorly estimated on shelves?}  So it seems possible the eddy has enough exchange with the surrounding water to have a shorter residence time.

There is partial mixing of the water in the eddy, though it remains vertically stratified.  The amount of turbulence required to homogenize 100 m of water over 90 days is $\kappa_{\rho} \approx 10^{-3}\ \mathrm{m^2\,s^{-1}}$.  Given that the diffusivity implied in the cross-channel surveys was very locally on the order of $\kappa_{\rho} = 10 \ \mathrm{m^2\,s^{-1}}$, this number is not outrageous if we think that such high dissipation is found in $10^{-4}$ of the water column.  If the diffusivity were higher, it could support a shorter residence time in the eddy, but there would need to be enough turbulence found on the shelf to support this.

The sharpness of the front with the eddy and the offshore water is also intriguing.  It was persistent for the duration of our detailed survey (\fref{fig:SpiceO2264}), and, so far as we can tell with the limited resolution, was present during the La Perouse cruises. There is not any substantial bathymetry blocking the onshore incursion of water at this location, so there must be a dynamic barrier. In contrast, numerical simulations of this region reported by \cite{sahuetal22} using a NEMO 36th-degree regional model indicate that there is relatively rapid exchange between the deep eddy water and the rest of the coastal ocean.  The region where the eddy resides has velocities equal or greater than other parts of the shelf, and water has an approximate residence time of less than 20 days.  Consequently, the eddy water does not develop a distinct $\theta$--$S$ signature from the surrounding water.

Overall, it would be an improvement to our understanding of the eddy if we could sample the shelf more persistently.  The eddy was already well-formed by the June La Perouse cruise, and seems to evolve slowly during that time.  Capturing its formation, presumably earlier in the spring, and its evolution through the year could answer

@@

The (apparent) displacement of shelf water at this location is relatively dramatic.  Eddies have been known to separate from irregular coastal topography, both at the surface \cite{barthetal00}, and deeper in the water column \cite{pellandetal13}.  However, the wholesale replacement of shelf water by a new water mass from offshore, as shown here is relatively rare. In our situation it is clear that water from as deep as 150 m is separating from the shelf and moving offshore (\fref{fig:SpiceO2264}, \fref{fig:IsopycnalsDepth}).

Satellite imagery makes it clear that there is often an exchange between coastal and deepwater along the Vancouver Island shelf (\fref{fig:SSTLateAug}).  Most years there are three of four large excursions from the shelf into the open ocean, many of them over 100 km long.  It is possible that there is even some spatial consistency to these features, with a persistent separation at the north tip of the Island, and a strong tendency for one at 49.5 N.  The study area also has evidence of a squirt in most of the years, with 2011 being the only clear exception.

>>>>

The southern Vancouver Island continental shelf is very productive due to high nutrient input from the Strait of Juan de Fuca and Salish Sea estuarine system, and substantial cross-shelf transport due to the complicated topography.  The estuarine water flows poleward, hugging the coast as the Vancouver Island Coastal Current, and in the summer shelf water flows equatorward.  Trapped between these two currents is the Juan de Fuca eddy, a stagnant region of water that occupies the shelf where it abruptly widens just north of the Strait of Juan de Fuca.

@@

Here we present detailed observations from the southern Vancouver Island shelf, collected in summer 2013, and sampled with a rapid profiling vehicle equipped with a CTD and oxygen sensor, supplemented by traditional hydrographic surveys. The shelf has complicated bathymetry \fref{fig:LocMapBoth}, with a somewhat typical shelf poleward of our study site, but then widening in the lee of La Perouse Bank, and then hitting the Juan de Fuca canyon, equatorward.  Water from the strait of Juan de Fuca flows poleward as a buoyant current that hugs the coast \cite{thomsonetal89, hickeyetal91}, while shelf water flows equatorward in the summer under upwelling conditions, both forced by local winds and via tele-connections with the long, homogenous shelves equatorward off Washington and Oregon \cite{hickeyetal91,thomsonkrassovski15,engidaetal16}.  Trapped between these two currents is a region of relatively homogenous water that has been termed the ``Juan de Fuca'' eddy \cite{freelanddenman82,freelandmcintosh89,foremanetal07,macfadyenhickey10}.

One of our findings below is that the water in the deep pool of the Juan de Fuca eddy is either  quite old or undergone significant vertical mixing.  The low oxygen found in the Juan de Fuca eddy has caused it to be inferred to be upwelled from as deep as 400 m \cite{freelanddenman82,deweycrawford88}.  This inference was based on the oxygen being conservative, and hence having to come from close to the oxygen minimum zone found further offshore \cite{mackasetal87}.  This lead to hypotheses that perhaps the eddy was fed by waters being drawn up a spur canyon of the Juan de Fuca canyon (called the Spur Canyon) via ageostrophic transport from offshore to onshore due to the low pressure in the eddy center \cite{weaverhsieh87}.  Below, we will argue that there is not substantial evidence of such transport, and that oxygen is likely low because of consumption due to respiration.

A second important finding is that the water flowing equatorward along the shelf from upstream all detaches from the shelf and is pushed offshore to be replaced by offshore water that floods the southern shelf.  Such events are important as they tend to transport nutrient rich shelf water out into the open ocean, fueling offshore ecosystems.  Similar offshore motion has been remarked in the region using satellite images \cite{ikedaemery84,thomsongower98}, and observed further to the south off Cape Blanco \cite{barthetal00}, and other coasts \cite[e.g.]{relvasbarton05}. Net cross-shore transport of nutrients and chlorophyll have been found in hydrographic surveys off Vancouver Island \cite{mackasyelland99}, and associated with mesoscale features both in geostrophic velocities, and in satellites.  The mechanisms driving such separations are poorly understood, but hypotheses include coastal hydraulics \cite{dalebarth01} leading to along-shore trapping of coastal waves,  wind stress curl variations \cite{castelaobarth07}, induced relative vorticity due to stretching of parcels that inertially overshoot their initial isobaths due to a sudden change in downstream bathymetry, and the non-linear breaking of large-scale meanders due to baroclinic instability between the wind-driven current and the California Undercurrent \cite{ikedaetal84, batteen97}.

In this paper we present  observations from 2013 (\fref{sec:Site}), presenting the data in a number of ways to highlight the important processes, with a focus on the offshore side of the eddy region (\fref{sec:Observations}) where we consider the age of the eddy, the steadiness of the front separating the eddy and the offshore water, properties along the spur canyon that was believed to feed the eddy, and finally describe the large offshore separation of the shelf-break current and its intrusion into the interior.  We discuss the origin of the eddy and the implications of the separating jet (\fref{sec:Summary}).

@@

\begin{figure*}[htbp]
  \includegraphics[width=5in]{LocMapBoth.pdf}
  \caption{a) Study site on the Vancouver Island Shelf.  The blue arrow indicates the direction of the Vancouver Island Coastal Current.  The pink arrow indicates southward flow of the coastal upwelling current.  The dashed box indicates the approximate limits of the study area.  b) The study area with hydrographic casts from the La Perouse cruises along the LB and LC Lines are green dots, hydrographic casts during the Falkor cruise indicated in yellow dots, and Moving Vessel Profiler casts as red dots.  The coordinate system used for this paper is shown with alternating grey and white bands at 10-km intervals.}
  \label{fig:LocMapBoth}
\end{figure*}

At the south end of the study site is the Juan de Fuca canyon, which feeds dense water into Juan de Fuca Strait.  This dense water is mixed with fresh water from the Fraser River at the sills and archipelagos further inland, and fluxes out the Strait again, where it turns poleward along Vancouver Island to form the Vancouver Island Coastal Current.  The Juan de Fuca Canyon has a notable spur canyon (Spur Canyon), that incises the shelf towards the north into the study site (\fref{fig:LocMapBoth}b).  The rest of the shelf is punctuated by a series of banks and shallow basins.  Notable is La Perouse bank, running along the shelf from the north, and moving isobaths offshore until it abruptly ends at around 48.5 N. The shelf break is characterized by a series of submarine canyons, in particular Nitnat Canyon and Barkley Canyon.

@@

Observations aboard the \emph{R/V Falkor} were carried out from 2013-08-21 to 2013-08-30, starting two weeks before the September \emph{Tully} cruise.  Data were collected with an ODIM Brooke-Ocean Moving Vessel Profiler, profiling to 200 m.  The MVP was equipped with an AML Oceanographic CTD, and a Rinko Oxygen sensor with a 7-s response time foil.  The MVP profiles at vertical fall speeds of $3\ \mathrm{m\,s^{-1}}$ to depths of 200 m or to within 5 m of the seafloor, whichever is shallower.  Data collection took place while the ship cruised at speeds between 5 and 8 kts, usually at around 6 kts to enable fine horizontal spacing of the casts, with typical spacing less than 800 m.  Data is collected during down- and up-casts, but is only reported for down-casts because the path through the water is highly asymmetric on the up-cast, whereas the downcast is largely vertical.  The rapid speed of the profiling makes the oxygen measurements somewhat coarse, and probably biased due to the phase lag of the sensor, so we treat these largely qualitatively in this paper.

@@

Winds during the cruises were typical for the west coast of Vancouver Island, with upwelling favorable winds during July and early August (\fref{fig:LaPeWind}).  During the cruise, and for the week previous, the winds were intermittently downwelling favorable.  Note that this locale is strongly affected by coastally trapped waves from further south, so doming of near-bottom isopycnals often persists despite local wind forcing \cite{thomsonkrassovski15, engidaetal16}, and takes a finite amount of time to spin down.

@@

Hydrographic sections along the LB and LC lines highlight summer conditions on the Southern Vancouver Island Shelf, and indicate some of the features we are focusing on in this paper (\fref{fig:LaPerouse2013Ctd}).  During both cruises and at both lines there is clear evidence of upwelling, with the $26.4\ \mathrm{kg\,m^{-3}}$ isopycnal reaching from 130 m to shallower than 85 m over the shelf.  This dense water tends to be low in oxygen and cool. Further onshore, isopycnals tilt back down, and warm water fills the water column, evidence of the Vancouver Island Coastal Current that hugs the coast here.

@@

The spice anomaly defined this way is mapped in the middle row of each set of plots in \fref{fig:LaPerouse2013Ctd}.  Offshore water tends to be high is absolute spice anomaly, with a positive-anomaly layer (red) centered at $26.4\ \mathrm{kg\,m^{-3}}$ sandwiched between negative-anomaly layers above and below.  On the shelf ($-10\, \mathrm{km} < X < 30\, \mathrm{km}$), the distinct $\theta$--$S$ masses are attenuated and much closer to the mixing line than the offshore water.  Hugging the coast ($X > 30\, \mathrm{km}$), is the Vancouver Island Coastal Current, with a negative spice anomaly in June, and a positive anomaly in summer, consistent with the warming of this water during the summer.

@@

A systematic survey through the front between the offshore water and the ``eddy'' water demonstrates the structure of the front (\fref{fig:Frontsurvey}).  The survey started close to shore and passed through the Vancouver Island Coastal Current (casts 2030--2080).  This water is buoyant and signficiantly fresher and cooler than water at the same density near the surface.  There is strong evidence of isopycnal mixing of the surface water and the outflow water, with warm-salty water in the relatively wide front between the ``eddy'' water and the water in the current.

@@

\subsection{Spur Canyon}

The seasurface is low in the middle of the ``Eddy'', so water has been hypothesized to move up the Spur Canyon due to ageostrophic motion \cite{weaverhsieh87,freelanddenman82}.  This is difficult to infer from water properties alone.  Three transects up the canyon indicate that deep isopycnals slope up into the canyon (\fref{fig:CanyonPropertiesSpice}, to approximately 30 km).  Across the eddy, isopycnals are largely flat until they intersect the Vancouver Island Coastal Current ($x\approx 80\ \mathrm{km}$).  The deeper isopycnals do not make it into the deeper basin northeast of La Perouse bank ($x\approx 70\ \mathrm{km}$).

@@

Much of the modified water in the Spur Canyon appears to come from the shelf to the west, but heavily modified by tidal moxing. A repeated tidal survey over the bank on the west side of the canyon shows a strong hydraulic response (\fref{fig:HydraulicsCanyon}) during onshore flow (17:58--21:00).  Note the tide here is largely diurnal, so this onslope flow only occurs once a day.  Given the stratification of $N\approx 6\times10^{-3} \ \mathrm{rad\,s^{-1}}$ and an overturning scale of 50 m, we might expect dissipation rates reaching $\epsilon \sim L^2N^{3} \approx 5\times10^{-4}\ \mathrm{m^2\,s^{-3}}$, which is more than three orders of magnitude higher than dissipation observed west of this location by \cite{deweycrawford88}.  The effect on the properties in the canyon is for water to spill over from the offshore front into the canyon during the tide.  This water is rapidly mixed with surrounding water such that its strong offshore spice values are attenuated.

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=4in]{HydraulicsCanyon}
    \caption{Repeated, tide-resolving survey across the Spur Canyon, time is indicated in the upper left of each plot (29 August, 2013).
      \label{fig:HydraulicsCanyon}
    }
  \end{center}
\end{figure*}

@@

We can attempt to track this feature in satellite imagery, though there are lots of clouds during this time period.  Sea-surface temperatures are generally colder coming out of the Strait of Juan de Fuca (\fref{fig:SST}).  Water flowing equatorward along the shelf tends to be cooler than offshore water, though this water is significantly saltier than the Juan de Fuca water hugging the shelf. On Aug-25, there is a cold tongue of water separating from La Perouse bank, and pointing south at 125.8 W with a cooler tendril streaming west at 48 N.  This feature is not as well-developed in the previous image, Aug 21, perhaps indicating that it is an evolving mesoscale feature.  By 31 Aug, there is no surface expression of the feature, though tendrils of cooler water can be seen separating from La Perouse Bank.  By Sep-05 the water gas significantly warmed, but the offshore pushing anomally does not appear to have a surface signature.

\begin{figure}
  \begin{center}
    \includegraphics[width=6in]{SSTSnaps}
    \caption{Sea surface temperature images from the observation period.  Grey areas are clouds; dashed gray line is the study area. Depths are contoured in thin gray lines at 200, 150 and 100 m. \cite{L2PMetopA}\label{fig:SST}}
  \end{center}
\end{figure}

Ccean chlorophyll estimates show the same feature (\fref{fig:ChlA}) demostrating the advection of a region of high chlorophyll to the west side of the eddy. It also shows a relatively high-chlorophyll squirt to the west, again exiting the study region at approximately 48 degrees N.  The feature is relatively long-lived, on the order of one month.  Inspection of images before 2013-08-05 were too obscured by clouds, or did not show this feature, and even by 2013-09-06 we see the feature fading from the satellite.  However, note that this feature is somewhat south of tongue that we observe deeper in the water column, but is indicative of substantial offshore transport in the region.

@@

Unfortunately, this data set cannot definitively track where the water in the eddy originated from.  The deepest water in the eddy could originate along the $\theta$--$S$ line from approximately 5.5 degrees to 7.5 degrees (\fref{fig:LaPerouse2013TS}a), which in the open ocean spans depths from 420 m to 70 m.  \citeA{mackasetal87} attempted to determine the origin of the water by including oxygen as a third variable to resolve the ambiguous $\theta$--$S$ relation.  However, as \fref{fig:LaPerouse2013TS}b makes clear, oxygen is definitely not a conserved property in the eddy, with concentrations up to $150 \ \mathrm{\mu mol\ kg^{1}}$ lower in the eddy than the water found offshore, and in a way that definitely cannot result from conservative mixing.  As a best guess, if the water in the deepest part of the eddy came from a vertical mixture of the water at $26.5\,\mathrm{kg\,m^{-3}}$ and an equal distance down in density space of $26.7\,\mathrm{kg\,m^{-3}}$, then the deepest water in the eddy would be coming from a depth of 250 m.  This is actually a relatively typical upwelling depth for coastal flows, and may not require extra input up Spur canyon as posited by \citeA{freelanddenman82}.

We found little evidence of vigorous flow up the Spur Canyon, or if there is, then the mixing in the canyon is quick enough to remove the $\theta$--$S$ signature of offshore water within 20 km (\fref{fig:CanyonPropertiesSpice}).  We did find substantial evidence of mixing in the canyon, but the primary route of high-spice water into the canyon appears to be due to tidal flow over the banks on the west side (\fref{fig:HydraulicsCanyon}), rather than evidence of flow up the canyon.  However, it is also of note that upwelling winds had ceased at this point (\fref{fig:LaPeWind}), meaning that the offshore pressure gradient may be reduced, leading to less ageostrophic upwelling in the canyon.  Whether the cessation of winds would also lead to a reduction of the low sea level height in the center of the eddy is an open question.

There is evidence of aging of the water in the eddy between late spring and late summer (\fref{fig:LaPerouse2013Ctd}), with a reduction of oxygen in the eddy.  If we posited that the reduction was all from the same water, then the oxygen consumption rate over the time between the late-May and early September cruises would be on the order of $0.5\ \mathrm{\mu mol\, kg^{-1} d^{-1}}$.  This seems on the low side of continental upwellng system estimates of apparent oxygen utilization rates, which are between $1$ and $5\  \mathrm{\mu mol\, kg^{-1} d^{-1}}$ \cite{dortchetal94}.   So it seems possible the eddy has enough exchange with the surrounding water to have a shorter residence time.

There is partial mixing of the water in the eddy, though it remains vertically stratified, and the amount of homogenization is such that either the mixing is very strong, or the water is retained in the eddy for a long time.  The amount of turbulence required to homogenize 100 m of water over 90 days is $\kappa \approx 10^{-3}\ \mathrm{m^2\,s^{-1}}$.  Given that the diffusivity implied in the cross-channel surveys was (very) locally on the order of $\kappa_{\rho} = 10 \ \mathrm{m^2\,s^{-1}}$, this number is not outrageous if we think that such high dissipation is found in $10^{-4}$ of the water column. We can more carefully quantify this by considering a synthetic profile or temperature and salinity based on an offshore profile, extrapolated from the bottom of the 200-m cast to 250 m, assuming that the temperature and salinity at 250 m are $6.2\ \mathrm{[^oC]}$ and $34\ \mathrm{psu}$ respectively (\fref{fig:TSExercise}).  Water at the offshore station was warmer than onshore, so the profile was also linearly interpolated to a surface value of ($31.6\ \mathrm{psu}$, $12\ \mathrm{^oC}$).  The profile was then linearly compressed into a depth of 150 m representative of the shelf depth in the dense pool, and subjected to mixing with a constant diffusivity of $\kappa = 10^{-3}\ \mathrm{m^2\,s^{-1}}$, with the surface and bottom values pinned under the assumption that the near-bottom source and surface waters are replenished from a large reservoir.  Similar to the naive scaling, the T-S relationship does not approach a straight line until after approximately 60-100 days, or until the mixing affects a vertical length scale of $\lambda = \left(\tau \kappa\right)^{1/2}$ that is greater than $\approx 70\,\mathrm{m}$.

\begin{figure*}[htbp]
  \begin{center}
    \includegraphics[width=4in]{TSExercise}
    \caption{Mixing model assuming constant eddy diffusivity of $\kappa = 10^{-3}\ \mathrm{m^2\,s^{-1}}$ acting on an offshore temperature and salinity profiles compressed from 250 thick to 150 m thick.  Profiles are pinned to the deep value at ($34\ \mathrm{psu}$, $6.2\ \mathrm{^oC}$), and a shallow one at ($31.6\ \mathrm{psu}$, $12\ \mathrm{^oC}$).
      \label{fig:TSExercise} }
  \end{center}
\end{figure*}

The sharpness of the front with the eddy and the offshore water is also intriguing.  It was persistent for the duration of our detailed survey (\fref{fig:SpiceO2264}), and, so far as we can tell with the limited resolution, was present during the La Perouse cruises. There is not any substantial bathymetry blocking the onshore incursion of water at this location, so there must be a dynamic barrier. In contrast, numerical simulations of this region reported by \cite{sahuetal22} using a NEMO 36th-degree regional model indicate that there is relatively rapid exchange between the deep eddy water and the rest of the coastal ocean.  In this model, the region where the eddy resides has velocities equal or greater than other parts of the shelf, and water has an approximate residence time of less than 20 days.  Consequently, the eddy water does not develop a distinct $\theta$--$S$ signature from the surrounding water.

Overall, it would be an improvement to our understanding of the eddy if we could sample the shelf more persistently.  The eddy was already well-formed by the June La Perouse cruise, and seems to evolve slowly during that time.  Capturing its formation, presumably earlier in the spring, and its evolution through the year would be valuable in undertanding retention and exchange on this productive part of the shelf.

@@

The (apparent) displacement of shelf water from La Perousse Bank is relatively dramatic.  Eddies have been known to separate from irregular coastal topography, both at the surface \cite{barthetal00}, and deeper in the water column \cite{pellandetal13}.   It has been recognized that instabilities lead to exchange between the open ocean and the shelf at this location \cite{ikedaemery84, ikedaemery84}.   However, observations of the wholesale replacement of shelf water by a new water mass from offshore, is relatively rare. In our situation it is clear that water from as deep as 150 m is separating from the shelf and moving offshore (\fref{fig:SpiceO2264}, \fref{fig:IsopycnalsDepth}).

Satellite imagery makes it clear that there is often an exchange between coastal and deepwater along the Vancouver Island shelf (\fref{fig:SSTLateAug}).  Most years there are three of four large excursions from the shelf into the open ocean, many of them over 100 km long.  The lengthscale in these images is somewhat larger than that inferred for this region by \citeA{ikedaetal84}.    It is possible that there is even some spatial locking of these features, with a persistent separation at the north tip of the Island, and a strong tendency for one at 49.5 N.  Our study area also has evidence of a squirt in most of the years, with 2011 being the only clear exception. General baroclinic instability of the upwelling front is a possible mechanism to drive offshore exchange \cite<e.g.>{ikedaetal84,durskiallen05}, but this tends to be shallow, with smaller-scale instabilities that will not extend as far into the interior ocean as observed here.  Rather it seems likely that the topographic change engendered by the sudden turn to the east of La Perouse bank catalyses a larger scale instability at this location.  In \citeA{durskiallen05}, when including realistic shelf bathymetry including Heceta Bank, the region around the bank catalyzes intermittent large-scale instabilities similar to the feature here and along the coast \cite{batteen97}.
